Stochastic Variance-Reduced Policy Gradient

### 2018-02-09 svrpg.tex
Updated
finished Appenddix B (Proofs)

\begin{definition}
For random vectors X, Y:
\begin{align*}
	\Covs{X}{Y} &\coloneqq \Tr\left(\Ets{(X-\Ets{X})(Y-\Ets{Y})^T}\right), \\
	\Vars{X} &\coloneqq \Covs{X}{Y},
\end{align*}
where $\Tr(\cdot)$ denotes the trace of a matrix. $\Covts[t]{X}{Y}$ and $\Varts[t]{X}$ are defined in the same way from $\Ets[t]{X}$.
\end{definition}

@@

For G(PO)MDP, we do not have a compact expression for $g$, but since it is derived from REINFORCE by neglecting some terms of the kind $\nabla \log \pi_{\vtheta}{a_t | s_t}\gamma^t r_t$ \cite{baxter2001infinite,peters2008reinforcement}, the above bound still holds. \todopir{we need to prove it formally if we have time}

@@

@@

where (\ref{eq:1}) is from the definition of $\gradApp{\vtheta}{N}$, (\ref{eq:2}) is from $\EVV[]{\norm[]{X-\EVV[]{X}}^2}\leq\EVV[]{\norm[]{X}^2}$ for any stochastic variable $X$\todopirout{ref}, (\ref{eq:3}) is from Lemma \ref{lemma:gsmooth}, 

@@

\todopir{Explain step $E[\|\|^2] \leq Var$}

@@

%Lemma 0
\begin{restatable}[]{lemma}{auxzero}\label{lemma:aux0}
Under Assumption \ref{ass:bounded_score}, for any function $\varphi(\vtheta_t^{s+1})$ which is deterministic for a fixed $\vtheta_t^{s+1}$:
\begin{align*}
\left|\Ets[t]{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
-\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
\right|
\leq
\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Ets[t-1]{\norm[]{\varphi(\vtheta_t^{s+1})}^2}
\end{align*}
\end{restatable}
\begin{proof}
\begin{align}
	\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
	&=
	\Ets{\dotprod{\gradIdeal{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{\varphi(\vtheta_t^{s+1})}} \label{eq:6}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{\varphi(\vtheta_t^{s+1})}} \label{eq:7}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\dotprod{\Ets{e_s}}{\Ets{\varphi(\vtheta_t^{s+1})}}
	+\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})}  \label{eq:new1}\\
	&= 
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})} \label{eq:8}
\end{align}
where~\eqref{eq:6} is from Definition~\ref{def:ideal};~\eqref{eq:7} is from the fact that $\gradIdeal{\vtheta_t^{s+1}}$ is both unbiased and independent from $\varphi(\vtheta_t^{s+1})$ \wrt the sampling at time $t$ alone, which is not true for $\gradBlack{\vtheta_t^{s+1}}$;~\eqref{eq:new1} is from the fact that $\gradJ{\wt{\vtheta}^s}$ is constant \wrt $\Vars{\cdot}$;~\eqref{eq:8} is from $\Ets{e_s}=0$.
Hence:
\begin{align}
	\left|\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
	\right.&-\left.\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}\right| 
	=
	\left|\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})}\right|  
	\nonumber\\
	&\leq
    \sqrt{\Vars{\gradApp{\wt{\vtheta}^s}{N}}} \cdot \sqrt{\Varts[t-1]{\varphi(\vtheta_t^{s+1})}} \label{eq:9a}\\
	%
	&\leq	
	\frac{1}{2}\Vars{\gradApp{\wt{\vtheta}^s}{N}} +\frac{1}{2}\Varts[t-1]{\varphi(\vtheta_t^{s+1})}\label{eq:9}\\
	%
	&=
	\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Varts[t-1]{\varphi(\vtheta_t^{s+1})} \label{eq:10}\\
	%
	&\leq
	\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Ets[t-1]{\norm[]{\varphi(\vtheta_t^{s+1})}^2},
	\nonumber
\end{align}
where~\eqref{eq:9a} comes from Cauchy-Schwarz inequality
\[
        |\Cov(X,Y)| = |\EVV[]{(X-\mu_X)\transpose{(Y-\mu_Y)}}| \leq \EVV[]{(X-\mu_X)^2}^{1/2}\EVV[]{(Y-\mu_Y)^2}^{1/2} = \sqrt{\Var(X) \Var(Y)},
\]
\eqref{eq:9} is from Young's inequality, and~\eqref{eq:10} is from the definition of $\gradApp{\vtheta}{N}$.
\todopir{comment the last step.}
\end{proof}

>>>>

\begin{definition}
For random vectors X, Y:
\begin{align*}
	\Covs{X}{Y} &\coloneqq \Tr\left(\Ets{(X-\Ets{X})(Y-\Ets{Y})^T}\right), \\
	\Vars{X} &\coloneqq \Covs{X}{Y},
\end{align*}
where $\Tr(\cdot)$ denotes the trace of a matrix. From the linearity of expected value we have the following:
\begin{equation}\label{neweq:1}
\Vars{X} = \Es{\norm[]{X-\Es{X}}^2} \leq \Es{\norm[]{X}^2}
\end{equation}
$\Covts[t]{X}{Y}$ and $\Varts[t]{X}$ are defined in the same way from $\Ets[t]{X}$.
\end{definition}

@@

For G(PO)MDP, we do not have a compact expression for $g$, but since it is derived from REINFORCE by neglecting some terms of the kind $\nabla \log \pi_{\vtheta}{a_t | s_t}\gamma^t r_t$ \cite{baxter2001infinite,peters2008reinforcement}, the above bound still holds.

@@

	+\Vars{\gradApp{\wt{\vtheta}^s}{N}} \nonumber\\
	\label{neweq:2}\\

@@

where (\ref{neweq:2}) is from (\ref{neweq:1}), (\ref{eq:1}) is from the definition of $\gradApp{\vtheta}{N}$, (\ref{eq:2}) is from (\ref{neweq:1}), (\ref{eq:3}) is from Lemma \ref{lemma:gsmooth}, 

@@

@@

%Lemma 0
\begin{restatable}[]{lemma}{auxzero}\label{lemma:aux0}
Under Assumption \ref{ass:bounded_score}, for any function $\varphi(\vtheta_t^{s+1})$ which is deterministic for a fixed $\vtheta_t^{s+1}$:
\begin{align*}
\left|\Ets[t]{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
-\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
\right|
\leq
\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Ets[t-1]{\norm[]{\varphi(\vtheta_t^{s+1})}^2}
\end{align*}
\end{restatable}
\begin{proof}
\begin{align}
	\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
	&=
	\Ets{\dotprod{\gradIdeal{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{\varphi(\vtheta_t^{s+1})}} \label{eq:6}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{\varphi(\vtheta_t^{s+1})}} \label{eq:7}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\dotprod{\Ets{e_s}}{\Ets{\varphi(\vtheta_t^{s+1})}}
	+\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})}  \label{eq:new1}\\
	&= 
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})} \label{eq:8}
\end{align}
where~\eqref{eq:6} is from Definition~\ref{def:ideal};~\eqref{eq:7} is from the fact that $\gradIdeal{\vtheta_t^{s+1}}$ is both unbiased and independent from $\varphi(\vtheta_t^{s+1})$ \wrt the sampling at time $t$ alone, which is not true for $\gradBlack{\vtheta_t^{s+1}}$;~\eqref{eq:new1} is from the fact that $\gradJ{\wt{\vtheta}^s}$ is constant \wrt $\Vars{\cdot}$;~\eqref{eq:8} is from $\Ets{e_s}=0$.
Hence:
\begin{align}
	\left|\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}
	\right.&-\left.\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{\varphi(\vtheta_t^{s+1})}}\right| 
	=
	\left|\Covts[t-1]{\gradApp{\wt{\vtheta}^s}{N}}{\varphi(\vtheta_t^{s+1})}\right|  
	\nonumber\\
	&\leq
    \sqrt{\Vars{\gradApp{\wt{\vtheta}^s}{N}}} \cdot \sqrt{\Varts[t-1]{\varphi(\vtheta_t^{s+1})}} \label{eq:9a}\\
	%
	&\leq	
	\frac{1}{2}\Vars{\gradApp{\wt{\vtheta}^s}{N}} +\frac{1}{2}\Varts[t-1]{\varphi(\vtheta_t^{s+1})}\label{eq:9}\\
	%
	&=
	\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Varts[t-1]{\varphi(\vtheta_t^{s+1})} \label{eq:10}\\
	%
	&\leq
	\frac{1}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} +\frac{1}{2}\Ets[t-1]{\norm[]{\varphi(\vtheta_t^{s+1})}^2},
	\label{neweq:3}
\end{align}
where~\eqref{eq:9a} comes from Cauchy-Schwarz inequality
\[
        |\Cov(X,Y)| = |\EVV[]{(X-\mu_X)\transpose{(Y-\mu_Y)}}| \leq \EVV[]{(X-\mu_X)^2}^{1/2}\EVV[]{(Y-\mu_Y)^2}^{1/2} = \sqrt{\Var(X) \Var(Y)},
\]
\eqref{eq:9} is from Young's inequality,~\eqref{eq:10} is from the definition of $\gradApp{\vtheta}{N}$, and \eqref{neweq:3} is from \eqref{neweq:1}.
\end{proof}
